---
title: "Getting started with ustats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ustats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Code chunks are not evaluated: the package requires a Python runtime
# with u-stats / numpy / torch, which is not available (and should not
# be downloaded) on the machines that build this vignette.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`ustats` is a thin R interface to the Python package
[`u-stats`](https://pypi.org/project/u-stats/), which computes
higher-order U-statistics efficiently using Einstein summation
(`numpy.einsum` / `torch.einsum`). This vignette covers the one part of
the package that needs a little care — **setting up the Python
environment** — and then shows basic usage.

## TL;DR

```{r}
install.packages("ustats")
library(ustats)

H <- matrix(rnorm(100), 10, 10)
ustat(list(H, H), "ab,bc->")
```

That is all most users need: on the first call, `reticulate` (>= 1.41)
automatically downloads a private Python together with the required
packages (`u-stats`, `numpy`, `torch`) into a cached environment, and
reuses it in later sessions.

## How the Python environment is resolved

`ustats` declares its Python dependencies via
`reticulate::py_require()` when the package is loaded. When Python is
first needed, reticulate resolves these requirements as follows:

1. If you have already configured a Python environment — via
   `reticulate::use_virtualenv()`, `reticulate::use_condaenv()`, or the
   `RETICULATE_PYTHON` environment variable — that environment is used.
   It must contain `u-stats`, `numpy`, and (recommended) `torch`.
2. Otherwise, reticulate provisions an **ephemeral, cached environment**
   containing the declared packages automatically. Nothing is installed
   into your system Python.

There are therefore three ways to set things up, from least to most
manual.

### Option 1: automatic (recommended)

Do nothing. The first call that touches Python triggers the automatic
setup:

```{r}
library(ustats)
ustat(list(matrix(rnorm(100), 10, 10)), "ab->")
```

Two things to know:

* The first call downloads Python packages once; later calls and later
  R sessions reuse the cache.
* On Linux, the default PyTorch build from PyPI bundles CUDA libraries
  (roughly 2.5 GB). If you do not have an NVIDIA GPU and prefer a small
  CPU-only build (roughly 200 MB), use Option 2 — or, before Python
  initializes, replace the declared requirement with a CPU-only one.

### Option 2: a persistent environment with `setup_ustats()`

`setup_ustats()` creates a dedicated environment and installs all
dependencies into it. By default it installs the **CPU-only** PyTorch
build:

```{r}
library(ustats)

setup_ustats()                # virtualenv/conda + CPU-only torch
setup_ustats(gpu = TRUE)      # default PyPI torch (CUDA-enabled on Linux)
setup_ustats(
  method  = "virtualenv",     # or "conda"
  envname = "r-ustats",
  persist = TRUE              # print the RETICULATE_PYTHON line to add
)                             # to your .Rprofile (no files are written)
```

For GPU builds on Windows, or for a wheel matching a specific CUDA
version, see <https://pytorch.org/get-started/locally/> and use
Option 3.

### Option 3: bring your own environment

If you already maintain a conda or virtualenv environment (for example,
one with a carefully chosen CUDA-enabled PyTorch), install the one
missing piece:

```bash
pip install u-stats
```

and tell reticulate to use that environment **before Python
initializes** (i.e. right after loading the package, before the first
`ustat()` call):

```{r}
library(ustats)
reticulate::use_condaenv("your_env_name", required = TRUE)
# or: reticulate::use_virtualenv("~/.virtualenvs/your_env")
```

Alternatively, set `RETICULATE_PYTHON` to the path of the Python binary
in `.Renviron` or `.Rprofile`, which takes effect for all sessions.

### Verifying the setup

```{r}
check_ustats_setup()
#> === ustats Environment Status ===
#>
#> [OK] Python: /path/to/python
#>   Version: 3.12
#> [OK] u_stats available
#> [OK] NumPy available
#> [OK] PyTorch available (version 2.5.1, CUDA available)
#>
#> ---------------------------------
#>  Environment fully ready (Torch backend available)
```

## Computing U-statistics

`ustat()` takes a list of kernel tensors (R vectors or matrices) and an
Einstein summation expression describing how their indices are
contracted, with distinct letters ranging over distinct observation
indices:

```{r}
library(ustats)

set.seed(1)
n <- 300

H1 <- rnorm(n)
H2 <- matrix(rnorm(n * n), n, n)
H3 <- rnorm(n)

result <- ustat(
  tensors    = list(H1, H2, H2, H3),
  expression = "a,ab,bc,c->",
  backend    = "torch",  # falls back to numpy if torch is unavailable
  average    = TRUE,     # divide by the number of index tuples
  dtype      = NULL      # auto: float32 on GPU, float64 on CPU
)
print(result)
```

The index structure can equivalently be given as a list of numeric
index vectors, which is convenient when the expression is built
programmatically:

```{r}
ustat(list(H1, H2, H2, H3), list(1, c(1, 2), c(2, 3), 3))
```

### GPU acceleration

With `backend = "torch"`, computations run on the GPU automatically
whenever PyTorch reports that CUDA is available:

```{r}
torch <- reticulate::import("torch")
torch$cuda$is_available()
```

## Troubleshooting

* **`check_ustats_setup()` reports a missing module.** The session is
  bound to a Python environment that lacks the dependency. Either
  install it there (`pip install u-stats`), or restart R and select a
  different environment (Options 2-3 above).
* **reticulate ignores `use_condaenv()` / `RETICULATE_PYTHON`.**
  reticulate binds to a single Python per R session, at the moment
  Python is first initialized. Restart R and configure the environment
  *before* anything touches Python.
* **The first call seems stuck.** It is most likely downloading
  PyTorch; see Options 1-2 for how to choose the much smaller CPU-only
  build.
* **`ustat()` warns "Torch backend not available".** The bound
  environment has no PyTorch; the computation falls back to NumPy,
  which is slower and can be less numerically stable. Install torch
  with `setup_ustats()` or
  `pip install torch --index-url https://download.pytorch.org/whl/cpu`.
